Enforced Invariant Confluence

### 2018-05-01 doc/paper/sections/introduction.tex
Honey, I shrunk the paper.

\section{Introduction}
When an application designer decides to replicate a piece of data, they have to
make a fundamental choice between weak and strong consistency. Replicating the
data with strong consistency using a technique like distributed transactions (e.g., ~\cite{bernstein1981concurrency,mohan1986transaction}) or state machine
replication (e.g., ~\cite{schneider1990implementing, lamport1998part,
liskov2012viewstamped, ongaro2014search}) makes the
application designer's life very easy. To the developer, a strongly
consistent system behaves exactly like a single-threaded system running on a
single node, so reasoning about the behavior of the system is
simple~\cite{herlihy1990linearizability}.  Unfortunately, strong consistency is
at odds with performance. The CAP theorem and PACELC theorem tell us that
strongly consistent systems suffer from higher latency at best and
unavailability at worst~\cite{gilbert2002brewer, brewer2012cap,
abadi2012consistency}. On the other hand, weak consistency models like eventual
consistency~\cite{vogels2009eventually}, PRAM
consistency~\cite{lipton1988pram}, causal consistency~\cite{ahamad1995causal},
and others~\cite{lloyd2011don, mehdi2017can} allow data to be replicated with
high availability and low latency, but they put a tremendous burden on the
application designer to reason about the complex interleavings of operations
that are allowed by these weak consistency models. In particular, weak consistency
models strip an application developer of one of the earliest and most effective
tools that is used to reason about the execution of programs: application
invariants~\cite{hoare1969axiomatic, balegas2015towards} such as database integrity constraints~\cite{IntegrityConstraintCitations}. Even if every
transaction executing in a weakly consistent system individually maintains an
application invariant, the system as a whole can produce invariant-violating
states.

Is it possible for us to have our strongly consistent cake and eat it with high
availability too? Can we replicate a piece of data with weak consistency but
still ensure that its invariants are maintained? Yes ... sometimes. Bailis et al.\
introduced the notion of \emph{\invariantconfluence{}} as a necessary and sufficient
condition for when invariants can be maintained over replicated data without
the need for any coordination~\cite{bailis2014coordination}. If an object is
\invariantconfluent{} with respect to an invariant, we can replicate it with
the performance benefits of weak consistency and (some of) the correctness
benefits of strong consistency.

@@

Designing such a general-purpose decision procedure is
difficult\footnote{In fact it's impossible. Determining
\invariantconfluence{} of an object is undecidable in general. Still, we can develop a
decision procedure that works well in the common case.}. Many prior
efforts instead develop a decision procedure for \emph{\invariantclosure{}}, a
sufficient (but not necessary) condition for
\invariantconfluence{}~\cite{li2012making, li2014automating}. The existing
approaches check whether an object is \invariantclosed{}. If it is, then they
conclude that it is also \invariantconfluent{}. If it's not, then the current
approaches are unable to conclude anything about whether or not the object is
\invariantconfluent{}. In this paper, we take a step back and study the
underlying reason \emph{why} \invariantclosure{} is not necessary for
\invariantconfluence{}. We find that \invariantconfluence{} is fundamentally a
reachability property, checking whether certain values of our replicated
object can actually be obtained in the execution of a system. We discover that
unreachable, yet invariant satisfying, states cause \invariantclosure{} to be
an unnecessary condition. Using this understanding, we construct a set of
modest conditions under which \invariantclosure{} and \invariantconfluence{}
are in fact \emph{equivalent}, allowing us to reduce the problem of determining
\invariantconfluence{} to that of determining \invariantclosure{}.

Then, we use these conditions to design a general-purpose interactive
\invariantconfluence{} decision procedure. Users iteratively interact with the
decision procedure, eliminating unreachable states from the invariant, moving
towards the conditions under which \invariantclosure{} and
\invariantconfluence{} are equivalent. Meanwhile, the decision procedure
generates examples of reachable and unreachable states which help the user
recognize patterns that describe reachability. We also leverage our intuitions
about reachability to develop a new sufficient condition for
\invariantconfluence{}, dubbed \emph{\mergereducibility}. \Mergereducibility{} covers
some cases that are not covered by \invariantclosure{}, and it can be
automatically checked without any user interaction.

@@

In closing, here is an outline of the paper and of our contributions:
\begin{itemize}
  \item
    We propose a novel expression-oriented definition of \invariantconfluence{}
    that is both formal and simple (\secref{InvariantConfluence}).

  \item
    We develop an understanding of why \invariantclosure{} is not necessary for
    \invariantconfluence{} and use this understanding to develop conditions
    under which it is both necessary and sufficient
    (\secref{InvariantClosure}).

  \item
    We exploit these conditions to design a general-purpose interactive
    decision procedure for \invariantconfluence{}
    (\secref{InteractiveDecisionProcedure}).

  \item
    We again exploit these conditions to design a novel non-trivial sufficient
    condition for \invariantconfluence{}, called \mergereducibility, that does
    not depend on user interaction (\secref{MergeReduction}).

  \item
    We present segmented \invariantconfluence{}: a generalization of
    \invariantconfluence{} that uses a small amount of coordination to maintain
    invariants for replicated objects that are otherwise not
    \invariantconfluent{} (\secref{SegmentedInvariantConfluence}).

  \item
    We implement our decision procedures and sufficient conditions for both
    \invariantconfluence{} and segmented \invariantconfluence{} in a system
    called Lucy, then use Lucy to evaluate our methods (\secref{Evaluation}).
\end{itemize}

>>>>

\section{Introduction}
When an application designer decides to replicate a piece of data, they have to
make a fundamental choice between weak and strong consistency. Replicating the
data with strong consistency using a technique like distributed transactions
(e.g., \cite{bernstein1981concurrency,mohan1986transaction}) or state machine
replication (e.g., \cite{schneider1990implementing, lamport1998part,
liskov2012viewstamped, ongaro2014search}) makes the application designer's life
very easy. To the developer, a strongly consistent system behaves exactly like
a single-threaded system running on a single node, so reasoning about the
behavior of the system is simple~\cite{herlihy1990linearizability}.
Unfortunately, strong consistency is at odds with performance. The CAP theorem
and PACELC theorem tell us that strongly consistent systems suffer from higher
latency at best and unavailability at worst~\cite{gilbert2002brewer,
brewer2012cap, abadi2012consistency}. On the other hand, weak consistency
models like eventual consistency~\cite{vogels2009eventually}, PRAM
consistency~\cite{lipton1988pram}, causal consistency~\cite{ahamad1995causal},
and others~\cite{lloyd2011don, mehdi2017can} allow data to be replicated with
high availability and low latency, but they put a tremendous burden on the
application designer to reason about the complex interleavings of operations
that are allowed by these weak consistency models. In particular, weak
consistency models strip an application developer of one of the earliest and
most effective tools that is used to reason about the execution of programs:
application invariants~\cite{hoare1969axiomatic, balegas2015towards} such as
database integrity constraints~\cite{grefen1993integrity, gupta1993local}. Even
if every transaction executing in a weakly consistent system individually
maintains an application invariant, the system as a whole can produce
invariant-violating states.

Is it possible for us to have our strongly consistent cake and eat it with high
availability too? Can we replicate a piece of data with weak consistency but
still ensure that its invariants are maintained? Yes... sometimes. Bailis et
al.\ introduced the notion of \emph{\invariantconfluence{}} as a necessary and
sufficient condition for when invariants can be maintained over replicated data
without the need for any coordination~\cite{bailis2014coordination}. If an
object is \invariantconfluent{} with respect to an invariant, we can replicate
it with the performance benefits of weak consistency and (some of) the
correctness benefits of strong consistency.

@@

Designing such a general-purpose decision procedure is difficult\footnote{In
fact it's impossible. Determining \invariantconfluence{} of an object is
undecidable in general. Still, we can develop a decision procedure that works
well in the common case.}. Many prior efforts instead develop a decision
procedure for \emph{\invariantclosure{}}, a sufficient (but not necessary)
condition for \invariantconfluence{}~\cite{li2012making, li2014automating}. The
existing approaches check whether an object is \invariantclosed{}. If it is,
then they conclude that it is also \invariantconfluent{}. If it's not, then the
current approaches are unable to conclude anything about whether or not the
object is \invariantconfluent{}.

In this paper, we take a step back and study the underlying reason \emph{why}
\invariantclosure{} is not necessary for \invariantconfluence{}.
%
% We find that \invariantconfluence{} is fundamentally a reachability property,
% checking whether certain values of our replicated object can actually be
% obtained in the execution of a system. We discover that unreachable, yet
% invariant satisfying, states cause \invariantclosure{} to be an unnecessary
% condition.
%
Using this understanding, we construct a set of modest conditions under which
\invariantclosure{} and \invariantconfluence{} are in fact \emph{equivalent},
allowing us to reduce the problem of determining \invariantconfluence{} to that
of determining \invariantclosure{}.
%
Then, we use these conditions to design a general-purpose interactive
\invariantconfluence{} decision procedure and a new sufficient condition for
\invariantconfluence{}, dubbed \emph{\mergereducibility}. \Mergereducibility{}
covers some cases that are not covered by \invariantclosure{}, and it can be
checked automatically.

% Then, we use these conditions to design a general-purpose interactive
% \invariantconfluence{} decision procedure. Users iteratively interact with the
% decision procedure, eliminating unreachable states from the invariant, moving
% towards the conditions under which \invariantclosure{} and
% \invariantconfluence{} are equivalent. Meanwhile, the decision procedure
% generates examples of reachable and unreachable states which help the user
% recognize patterns that describe reachability. We also leverage our intuitions
% about reachability to develop a new sufficient condition for
% \invariantconfluence{}, dubbed \emph{\mergereducibility}. \Mergereducibility{}
% covers some cases that are not covered by \invariantclosure{}, and it can be
% automatically checked without any user interaction.

@@

% TODO: add bit on performance.

In closing, here is an outline of the paper and of our contributions:
%
We propose a novel expression-oriented definition of \invariantconfluence{}
that is both formal and simple (\secref{InvariantConfluence}).
%
We develop an understanding of why \invariantclosure{} is not necessary for
\invariantconfluence{} and use this understanding to develop conditions under
which it is both necessary and sufficient (\secref{InvariantClosure}).
%
We exploit these conditions to design a general-purpose interactive decision
procedure for \invariantconfluence{} (\secref{InteractiveDecisionProcedure}).
%
We again exploit these conditions to design a novel non-trivial sufficient
condition for \invariantconfluence{}, called \mergereducibility.
%
We present segmented \invariantconfluence{}: a generalization of
\invariantconfluence{} that uses a small amount of coordination to maintain
invariants for replicated objects that are otherwise not \invariantconfluent{}
(\secref{SegmentedInvariantConfluence}).
%
We implement our decision procedures and sufficient conditions for both
\invariantconfluence{} and segmented \invariantconfluence{} in a system called
Lucy, then use Lucy to evaluate our methods (\secref{Evaluation}).

### 2018-05-01 doc/paper/sections/related_work.tex
Honey, I shrunk the paper.

\textbf{The Demarcation and Homeostasis Protocols.}
The homeostasis protocol~\cite{roy2015homeostasis}, a generalization of the
demarcation protocol~\cite{barbara1994demarcation}, uses program analysis to
avoid unnecessary coordination between servers in a \emph{sharded} database
(whereas \invariantconfluence{} targets \emph{replicated} databases).  The
protocol guarantees that transactions are executed with observational
equivalence with respect to some serial execution of the transactions. This
means that intermediate states may be inconsistent, but externally observable
side effects and the final database state are consistent.  The observational
equivalence guaranteed by the homeostasis protocol is stronger than the
guarantees of \invariantconfluence{}. As a result, there are invariants and
workloads that the homeostasis protocol would execute with more coordination
than a segmented \invariantconfluent{} execution.  The protocol generates
global invariants, called global treaties, and servers operate without
coordination so long as they preserve the treaty. When the treaty is violated,
the nodes perform a round of coordination and establish a new treaty.  This
mechanism of establishing invariants and operating without coordination so long
as the invariants are maintained is very similar to our design of segmented
\invariantconfluence{}.

>>>>

\textbf{The Demarcation and Homeostasis Protocols.}
The homeostasis protocol~\cite{roy2015homeostasis}, a generalization of the
demarcation protocol~\cite{barbara1994demarcation}, uses program analysis to
avoid unnecessary coordination between servers in a \emph{sharded} database
(whereas \invariantconfluence{} targets \emph{replicated} databases).  The
protocol guarantees that transactions are executed with observational
equivalence with respect to some serial execution of the transactions. This
means that intermediate states may be inconsistent, but externally observable
side effects and the final database state are consistent.  The observational
equivalence guaranteed by the homeostasis protocol is stronger than the
guarantees of \invariantconfluence{}. As a result, there are invariants and
workloads that the homeostasis protocol would execute with more coordination
than a segmented \invariantconfluent{} execution. Moreover, the homeostasis and
demarcation protocols' mechanism of establishing global invariants and
operating without coordination so long as the invariants are maintained is very
similar to our design of segmented \invariantconfluence{}.

### 2018-05-01 doc/paper/sections/segmented_invariant_confluence.tex
Honey, I shrunk the paper.

\begin{itemize}
  \item
    If $t \notin T_i$, then $p_i$ initiates a round of global coordination to
    execute $t$. During global coordination, every server temporarily stops
    processing transactions and transitions to state $s = s_1 \join \ldots
    \join s_n$, the join of every server's state. Then, every server
    speculatively executes $t$ transitioning to state $t(s)$. If $t(s)$
    violates the invariant (i.e.  $\lnot I(t(s))$), then every server aborts
    $t$ and reverts to state $s$.  Then, $p_i$ replies to the client. If $t(s)$
    satisfies the invariant (i.e.  $I(t(s))$), then every server commits $t$
    and remains in state $t(s)$. Every server then chooses the first segment
    $(I_i, T_i) \in \Sigma$ such that $I_i(t(s))$ to be the new active segment.
    Note that such a segment is guaranteed to exist because the segment
    invariants cover $I$. Moreover, $\Sigma$ is ordered, so every server will
    deterministically pick the same active segment. In fact, an invariant of
    the system model is that at any given point of normal processing, every
    server has the same active segment.

  \item
    If $t \in T_i$, then $p_i$ executes $t$ immediately and without
    coordination.  If $t(s_i)$ satisfies the \emph{active} invariant (i.e.
    $I_i(t(s_i))$), then $p_i$ commits $t$, stays in state $t(s_i)$, and
    replies to the client. If $t(s_i)$ violates the \emph{global} invariant
    (i.e. $\lnot I(t(s_i))$), then $p_i$ aborts $t$, reverts to state $s_i$,
    and replies to the client. If $t(s_i)$ satisfies the global invariant but
    violates the active invariant (i.e.  $I(t(s_i))$ but $\lnot I_i(t(s_i))$),
    then $p_i$ reverts to state $s_i$ and initiates a round of global
    coordination to execute $t$, as described in the previous paragraph.
    Transaction execution is summarized in \algoref{TxnExecution}.
\end{itemize}

@@

Second, segmented \invariantconfluence{} naturally subsumes a distributed
locking approach to replicating non-\invariantconfluent{} objects.  This
approach first recognizes which transactions cannot be safely executed
concurrently and then requires them to acquire a distributed lock before
executing~\cite{balegas2015putting, gotsman2016cause}. For example, in a
banking application with the invariant that all balances remain non-negative,
concurrent deposits are permitted, but concurrent withdrawals can lead to
invariant violations. Thus, we require that withdrawals acquire a distributed
lock before executing. This example is exactly the same as
\exampleref{CounreachableExample} which we handled by simply removing
withdrawal transactions from our segmentation's set of transactions.

Third, we can integrate a couple of optimizations into our system model to
further reduce the amount of coordination it requires. To begin, if a server with
state $s_i$ and active segment $(I_i, T_i)$ receives a transaction $t \in I_i$
to execute, and $t(s_i)$ violates the active invariant but not the global
invariant, instead of initiating a round of global coordination, $p_i$ can
simply buffer $t$ for re-execution at a later time. While this increases the
latency required to execute $t$, it's possible that after other transactions
are executed, re-executing $t$ may lead to a state that either satisfies the
active invariant or violates the global invariant. In either case, a round of
global coordination is avoided. Similarly, servers can buffer transactions that
require global coordination, executing an entire batch of these transactions
during a single round of global coordination.

Fourth, a segmented \invariantconfluence{} decision procedure can also leverage
\thmref{LatticeProperty} in addition to
\algoref{ArbitraryStartInteractiveDecisionProcedure}. If an object $O$ meets
criteria (1) - (3), then it is \sticonfluent{s}{T}{I} for every state $s \in
I$.

Fifth, while a segmented \invariantconfluence{} decision procedure can help
decide whether or not an object is segmented \invariantconfluent{}, it cannot
currently help construct a segmentation. In the future, we plan to explore
ways by which we can automatically suggest segmentations to the application
designer.

>>>>

If $t \notin T_i$, then $p_i$ initiates a round of global coordination to
execute $t$. During global coordination, every server temporarily stops
processing transactions and transitions to state $s = s_1 \join \ldots \join
s_n$, the join of every server's state. Then, every server speculatively
executes $t$ transitioning to state $t(s)$. If $t(s)$ violates the invariant
(i.e.  $\lnot I(t(s))$), then every server aborts $t$ and reverts to state $s$.
Then, $p_i$ replies to the client. If $t(s)$ satisfies the invariant (i.e.
$I(t(s))$), then every server commits $t$ and remains in state $t(s)$. Every
server then chooses the first segment $(I_i, T_i) \in \Sigma$ such that
$I_i(t(s))$ to be the new active segment.  Note that such a segment is
guaranteed to exist because the segment invariants cover $I$. Moreover,
$\Sigma$ is ordered, so every server will deterministically pick the same
active segment. In fact, an invariant of the system model is that at any given
point of normal processing, every server has the same active segment.

Otherwise, if $t \in T_i$, then $p_i$ executes $t$ immediately and without
coordination.  If $t(s_i)$ satisfies the \emph{active} invariant (i.e.
$I_i(t(s_i))$), then $p_i$ commits $t$, stays in state $t(s_i)$, and replies to
the client. If $t(s_i)$ violates the \emph{global} invariant (i.e. $\lnot
I(t(s_i))$), then $p_i$ aborts $t$, reverts to state $s_i$, and replies to the
client. If $t(s_i)$ satisfies the global invariant but violates the active
invariant (i.e.  $I(t(s_i))$ but $\lnot I_i(t(s_i))$), then $p_i$ reverts to
state $s_i$ and initiates a round of global coordination to execute $t$, as
described in the previous paragraph.  Transaction execution is summarized in
\algoref{TxnExecution}.

@@

% Second, segmented \invariantconfluence{} naturally subsumes a distributed
% locking approach to replicating non-\invariantconfluent{} objects.  This
% approach first recognizes which transactions cannot be safely executed
% concurrently and then requires them to acquire a distributed lock before
% executing~\cite{balegas2015putting, gotsman2016cause}. For example, in a
% banking application with the invariant that all balances remain non-negative,
% concurrent deposits are permitted, but concurrent withdrawals can lead to
% invariant violations. Thus, we require that withdrawals acquire a distributed
% lock before executing. This example is exactly the same as
% \exampleref{CounreachableExample} which we handled by simply removing
% withdrawal transactions from our segmentation's set of transactions.

% Second, we can integrate a couple of optimizations into our system model to
% further reduce the amount of coordination it requires. To begin, if a server with
% state $s_i$ and active segment $(I_i, T_i)$ receives a transaction $t \in I_i$
% to execute, and $t(s_i)$ violates the active invariant but not the global
% invariant, instead of initiating a round of global coordination, $p_i$ can
% simply buffer $t$ for re-execution at a later time. While this increases the
% latency required to execute $t$, it's possible that after other transactions
% are executed, re-executing $t$ may lead to a state that either satisfies the
% active invariant or violates the global invariant. In either case, a round of
% global coordination is avoided. Similarly, servers can buffer transactions that
% require global coordination, executing an entire batch of these transactions
% during a single round of global coordination.

Second, we can integrate a couple of optimizations into our system model to
further reduce the amount of coordination it requires. For example, servers can
buffer transactions that require global coordination, executing an entire batch
of these transactions during a single round of global coordination.

Third, a segmented \invariantconfluence{} decision procedure can also leverage
\thmref{LatticeProperty} in addition to
\algoref{ArbitraryStartInteractiveDecisionProcedure}. If an object $O$ meets
criteria (1) - (3), then it is \sticonfluent{s}{T}{I} for every state $s \in
I$.

Fourth, while a segmented \invariantconfluence{} decision procedure can help
decide whether or not an object is segmented \invariantconfluent{}, it cannot
currently help construct a segmentation. In the future, we plan to explore
ways by which we can automatically suggest segmentations to the application
designer.
